---
title: "Case Study 1"
author: "Jonathan Sneh, Ishani Tarafdar, Georges Durand, Raul Higareda"
date: "2023-03-28"
output: pdf_document
---


## Data Explorations and Summary Statistics

```{r}
grades <- read.csv("grades.csv", header=TRUE)
dim(grades)
```


## Model Selection

We want to make a model with 95% confidence (i.e. $\alpha$ )

```{r}
grades.mlr <- lm(exam2 ~ . ,data=grades)
summary(grades.mlr)
```

Based on the full model summary above, we may want to look into dropping project, cs, and participation from the dataset 
since the p-values in the summary output for project, cs, and participation are all greater than 0.05. 



Our null and alternative hypothesis are as follows: 
\[
\begin{cases}
    H_0,& \beta_{project} = \beta_{cs} = \beta_{participation}\\
    H_A, & \text{Either } \beta_{project}, \beta_{cs}, \text{ or } \beta_{participation} \text{ is not equal to zero}
\end{cases}
\]


```{r}
grades.reducedmlr = lm(exam2~exam1 + semester + hw, data=grades)
summary(grades.reducedmlr)
anova(grades.mlr, grades.reducedmlr)
```
Since the p-value is greater than an alpha level of 0.05 in the anova ouput, we fail to reject the null hypothesis with 95% level of confidence.



## Unusual Observations and Model Assumptions